---
title: "polyreg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{polyreg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE, warning=FALSE}
library(devtools)
devtools::load_all()
install_github("gestimation/polyreg") 
#pak::pak("gestimation/polyreg")
#devtools::install_github("gestimation/polyreg") 
library(polyreg)
#library(mets)
library(nleqslv)is
library(modelsummary)
```

## Overview

The polyreg package implements direct polynomial regression, a model that estimates multiplicative effects (e.g., risk ratio, odds ratio, or sub-distribution hazard ratio) on risks, cumulative incidence probabilities at a specific time point, or common effects over time. A key feature of this model is its ability to analyze multiple competing events simultaneously while ensuring that the probabilities sum to one. This is achieved by reparameterizing nuisance parameters using polytomous log odds products. Additionally, the package supports direct binomial regression for survival outcomes and the Richardson model for binomial outcomes, both of which use log odds products.

The models in polyreg are specified by three main components:

Nuisance model: Describes the relationship between outcomes and covariates (excluding exposure).

Effect measures and time points: Defines the exposure effect to be estimated and the time point of interest.

Censoring adjustment: Specifies strata for inverse probability weighting to adjust for dependent censoring.

## Model specification
### 1. Nuisance Model

The nuisance.model argument specifies the formula linking the outcome to covariates. Its format depends on the outcome type:

Competing risks or survival outcome: Use Surv() or Event() with time and status variables. 

Binomial outcome: Use standard R formula notation.

Default event codes:

Competing risks outcome: 1 and 2 for event types, 0 for censored observations.

Survival outcome: 1 for events, 0 for censored observations.

Binomial outcome: 0 and 1.

Event codes can be customized using code.event1, code.event2, and code.censoring. The outcome.type argument must be set to:

Effects on cumulative incidence probabilities at a specific time: 'COMPETINGRISK'

Effects on a risk at a specific time: 'SURVIVAL'

Effects on a risk of a binomial outcome: 'BINOMIAL'

Common effects on cumulative incidence probabilities over time: 'PROPORTIONAL'

Covariates included in nuisance.model should adjust for confounding factors to obtain unbiased exposure effect estimates. 

### 2. Effect measures and time points

Three effect measures available:

Risk Ratio (RR)

Odds Ratio (OR)

Sub-distribution Hazard Ratio (SHR)

Set the desired measure using effect.measure1 and, for competing risks analysis, effect.measure2. The time.point argument specifies the follow-up time at which effects are estimated. 

### 3. Censoring adjustment

Inverse probability weights adjust for dependent censoring. Use the strata argument to specify stratification variables. If no strata are specified, Kaplan-Meier weights are used. 


## Output

The main components of the output list include:

coefficient: Regression coefficients

cov: Variance-covariance matrix

diagnosis.statistics: Inverse probability weights, influence functions, and predicted values 

summary: Summary of exposure effect estimates

summary.full: Detailed summary including all regression coefficients

Use the summary or summary.full output with msummary() to display formatted results. The regression coefficients and their variance-covariance matrix are provided as coefficient and cov, respectively, with the first element corresponding to the intercept term, subsequent elements to the covariates in nuisance.model, and the last element to exposure. Finally, diagnosis.statistics is a dataset containing inverse probability weights, influence functions, and predicted values of the potential outcomes of individual observations.

## Example 1. Unadjusted competing risks analysis

For the initial illustration, unadjusted analysis focusing on cumulative incidence probabilities of event 1 and 2 at 8 years is demonstrated. Regression coefficients and variance covariance matrix of both exposure (fruitq1) and covariates (intercept in this case) in the fitted direct polytomous regression are presented. 

```{r example1}
library(polyreg)
data(diabetes.complications)
output <- polyreg(nuisance.model = Event(t,epsilon) ~ 1, exposure = 'fruitq1', data = diabetes.complications,
          effect.measure1='RR', effect.measure2='RR', time.point=8, outcome.type='C')
print(output$coefficient)
print(output$cov)
```

The summaries of analysis results in the list of outputs (e.g. output$summary.full below) are in accordance with the format of model summary function. All regression coefficients above are included in summary.full and model summary may be used to converted to risk ratios, odds ratios or sub-distribution hazards ratios by selecting exponentiate=TRUE option. The summaries can be displayed in Viewer with customized statistics such as p-values or confidence intervals. 

```{r example2}
msummary(output$summary.full, statistic = c("conf.int"), exponentiate = TRUE)
```

## Example 2. Survival analysis

The second example is survival analysis (outcome.type='SURVIVAL') to estimate the effects on the risk of diabetic retinopathy at 8 years of follow-up, treating macrovascular complications as censoring. 15 covariates and censoring strata are specified in nuisance.model= and strata=, respectively.

```{r example3}
data(diabetes.complications)
diabetes.complications$d <- (diabetes.complications$epsilon>0)
output <- polyreg(nuisance.model = Event(t,d) ~ age+sex+bmi+hba1c+diabetes_duration
          +drug_oha+drug_insulin+sbp+ldl+hdl+tg+current_smoker+alcohol_drinker+ltpa, 
          exposure = 'fruitq1', strata='strata', data = diabetes.complications,
          effect.measure1='RR', time.point=8, outcome.type='SURVIVAL')
```

Only the regression coefficient of exposure is included in summary, so now model summary does not display parameters of the covariates other than exposure. 

```{r example4}
msummary(output$summary, statistic = c("conf.int"), exponentiate = TRUE)
```

## Example 3. Binomial analysis

Binomial analysis is conducted if outcome.type='BINOMIAL'. Outcomes of observations censored before 8 years of follow-up are now treated as complication-free. 

```{r example5}
output <- polyreg(nuisance.model = d ~ age+sex+bmi+hba1c+diabetes_duration
          +drug_oha+drug_insulin+sbp+ldl+hdl+tg+current_smoker+alcohol_drinker+ltpa, 
          exposure = 'fruitq1', strata='strata', data = diabetes.complications,
          effect.measure1='RR', time.point=8, outcome.type='BINOMIAL')
msummary(output$summary, statistic = c("conf.int"), exponentiate = TRUE)
```

## Example 4. Competing risks analysis

The code below specifies direct polytomous regression of both of competing events (outcome.type='COMPETINGRISK'). Initial values of regression parameters are imported by data.initial.values. 

```{r example6}
data.initial.values <- c(-20.01, 1.591, -0.2065, 1.962, 10.05, 4.004, 0.9624, 1.582, 2.414, -0.02140,
                         1.072, 1.304, -0.6444, -1.600, -0.01816, 0.4460, -29.25, 6.586, -1.090, 3.515,
                         4.048, -5.012, 0.05348, 0.1609, 6.432, -0.4783, -3.230, -4.135, 1.451, 0.04964,
                         3.822, -0.08824)
output <- polyreg(nuisance.model = Event(t,epsilon) ~ age+sex+bmi+hba1c+diabetes_duration
          +drug_oha+drug_insulin+sbp+ldl+hdl+tg+current_smoker+alcohol_drinker+ltpa, 
          exposure = 'fruitq1', strata='strata', data=diabetes.complications,
          effect.measure1='RR', time.point=8, outcome.type='COMPETINGRISK',
          data.initial.values=data.initial.values)
msummary(output$summary, statistic = c("conf.int"), exponentiate = TRUE)
```
